Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-12 manuscript/revision2/revision2.tex
revision2: generate revised figures 1-4 and SI figures 1-4

Narrow-sense heritabilities (h\textsuperscript{2}) suggested that
rank-changing GxE for these phenotypes was present across the common
gardens (Figure~\ref{fig-map} C). h\textsuperscript{2} were typically
high at individual gardens: 59\% on average for green-up date, and 87\%
for flowering date. However, h\textsuperscript{2} were variable across
gardens, and green-up dates were uncorrelated (r\textsuperscript{2}
\textless{} 0.2) or negatively correlated between pairs of gardens
(Figure~\ref{fig-map} B). These negative and small correlations
undoubtedly contributed to the low h\textsuperscript{2} values for
green-up and flowering date when estimated jointly at all eight gardens:
h\textsuperscript{2} was 0.8\% for green-up and 23.2\% for flowering
date.

@@

We expected patterns of GxE and GxWeather to vary at the locus,
subpopulation, and trait level; thus we were interested in modeling the
different ways effects might covary across our eight common gardens. We
jointly re-estimated genetic effects of a subset of SNPs across all
eight common gardens using \emph{mash} models, which can flexibly
capture many modes of covariance across gardens (Figure~\ref{fig-map} D;
SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects,
we used subsets of effects from garden- and subpopulation GWAS (SI
Appendix, Section S2-S3; Table S1). To obtain covariance structures, we
specified both covariance structures introduced in the initial
\emph{mash} manuscript, and GxWeather covariance matrices estimated from
the covariance of empirical weather patterns at each garden at specific
times before the phenological event (Table 1; SI Appendix, Section S1).
Then, we used a greedy algorithm to select covariance matrices from each
model's set that significantly improved the model likelihood (SI
Appendix, Section S4; Table S2). Finally, we fit \emph{mash} models
using the selected set of covariance matrices on a subset relatively
unlinked (r\textsuperscript{2} \textless{} 0.2), `strong' genetic
effects with low \emph{p}-values (SI Appendix, Section S5).

@@

Eight GxWeather covariance structures were selected by the greedy
algorithm, one to two per subpopulation: five for green-up date, and
three for flowering date (SI Appendix, Table S2). Of these eight, five
had mass on them in \emph{mash} models of the strong effects
(Figure~\ref{fig-covar} A-B). Three of these five matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while two had all positive or near-zero
covariances.

@@

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited rank-changing GxE, or a difference in sign
between pairs of Texas and North gardens (Figure~\ref{fig-effects} B; SI
Appendix, Fig S1). 78.7\% of pairwise comparisons between North and
Texas gardens had a difference in sign, while only 28.6\% and 0.2\% of
North-North or Texas-Texas comparisons had a difference in sign,
respectively (Figure~\ref{fig-effects} B). The majority of pairwise
effects for greenup for the Midwest (\textgreater55\%) and Both
(\textgreater85\%) subpopulations were the same sign, and effects most
often differed in magnitude between gardens within and between the
regions (Figure~\ref{fig-effects} B).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited rank-changing GxE, or a difference in effect sign
within or between regions (Figure~\ref{fig-effects} B). More effects
differed in magnitude between the Texas and North regions than within
these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects} B),
while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering, but a large proportion
of these differed in sign between Texas and North regions (42.7\%) or
within the North region (65.4\%). Finally, in Both subpopulations, less
than 20\% of pairwise effects differed in sign (Figure~\ref{fig-effects}
B). Most differences in sign were between TX1, the southernmost garden,
and all other gardens (SI Appendix, Figure S1). Similarly, more effects
that differed in magnitude included gardens in the Texas region
(52.3-55.9\%), and most effect pairs in the North region were not
distinguishable (91.5\%).

@@

We sought additional experimental support the occurrence and location of
our significant SNP effect re-estimates using an independent pseudo-F2
mapping population created from Gulf \& Midwest individuals and grown at
the same sites (Figure~\ref{fig-qtl} A,B). We conducted quantitative
trait loci (QTL) mapping of flowering as functions of five environmental
cues that we also used as covariance matrices in \emph{mash}, and
identified eight QTL for flowering date, eight QTL for flowering as a
function of day length change two days prior, one QTL for the start of
vegetative growth, and two QTL for vegetative growth as a function of
daylength change one day prior, all of which showed QTL by environment
interactions (SI Appendix, Figure S2). All QTL for flowering overlapped
one or more homologs from rice or \emph{A. thaliana} with functionally
validated roles in flowering (SI Appendix, Dataset 7); the QTL on Chr04K
overlapped a gene recently functionally validated for flowering in
switchgrass (38). All flowering and green-up QTL intervals contained at
least one SNP significant in at least one \emph{mash} run at a
log10-transformed Bayes Factor \textgreater{} 2, or in the 1\% tail of
significance, whichever was stricter (SI Appendix, Dataset 8). We also
looked for enrichments of \emph{mash} SNPs in the 1\% tail of
significance (the `\emph{mash} 1\% tail') within each QTL interval. At
the 5\% level, three QTL had enrichments of SNPs in the \emph{mash} 1\%
tail. Overall, there were five significant enrichments (p \textless{}
0.05, hypergeometric test) of SNPs in the \emph{mash} 1\% tail in the
QTL intervals. Thus, we were able to experimentally support the genomic
windows of some re-estimates of significant SNP effects with a QTL
mapping experiment using a separate mapping population.

@@

\includegraphics{images/Fig1_200dpi.png}

@@

\caption{\label{fig-map}Characterization of the timing of vegetative
(green-up) and reproductive (flowering) growth in the switchgrass
diversity panel. (a) Map and trait histograms of green-up and flowering
dates across two genetically distinct switchgrass subpopulations and
eight common gardens. Purple represents individuals from the Midwest
genetic subpopulation, and pink individuals from the Gulf subpopulation;
map positions represent the original collection locations for the
genotypes, and shapes represent the ecotype of the genotype. Vertical
dashed lines indicate the summer solstice. Common gardens are arranged
in latitudinal order. (b) Phenotypic correlations between clonal
replicates planted at eight common gardens, within and between two
genetic subpopulations. (c) Narrow sense heritability of green-up and
flowering within single common gardens (purple) and across all eight
common gardens (green), within and between two genetic subpopulations.
(d) Flow diagram of the methods applied to the green-up and flowering
dates to jointly estimate SNP effects across all sites. \emph{Mash} was
fit to SNP effect data and used to find covariance matrices that
improved the \emph{mash} model likelihood using a large set of randomly
selected, relatively unlinked SNP effects; this model was applied to a
``strong'' set of SNP effects with large effect sizes in the univariate
GWAS.}

@@

\includegraphics{images/Figure_2_recolor_covariance_matrices_and_posterior_weights.png}

@@

\caption{\label{fig-covar}Example hypothesis-driven covariance matrices
specified in \emph{mash} and the posterior weights placed on all
covariance matrices. (a) Left column: Five example hypothesized
covariance matrices specified for the green-up date or flowering date
phenotype; these matrices were created from environment-specific
correlations across eight common gardens, and are described in Table 1.
Common gardens are arranged in latitudinal order within the matrices.
Right column: Five example canonical covariance matrices. Canonical
matrices (purple) have simple interpretations, such as equal effects
across all common gardens, or effects specific to a single common
garden. (b,d) Total posterior weight placed on each covariance matrix
type specified for (b) green-up date and (d) flowering date \emph{mash}
models, within and between two genetic subpopulations. Hypothesized
covariance matrices (green). Covariance matrices included in \emph{mash}
that had zero posterior weight in all three \emph{mash} runs on the
genetic subpopulations, such as the identity matrix, are not shown.
(c,e) Total posterior weight placed on covariance matrices that were
hypothesized or canonical, for the (c) green-up date phenotype and (e)
flowering date phenotype.}

@@

\caption{\label{fig-effects}Types of GxE present between of pairs of
jointly re-estimated effects in eight common gardens, for effects with
lfsr \textless{} 0.05. a) Examples of effect patterns at three pairs of
sites with three types of GxE. All effects are for an alternate allele,
with the reference allele effect defined to be zero at both gardens.
Sign: Effects that differ in sign at these pairs of gardens (p
\textless{} 0.05, lfsr). Magnitude: Effects identical in sign (p
\textless{} 0.05, lfsr) that differ in magnitude by a factor of
\textgreater0.4. None: Effects not distinguishable by magnitude nor sign
of the effect. b) The fraction of effects with each GxE type for the
start of vegetative growth (Greenup Date) and reproductive growth
(Flowering Date), within and between two genetic subpopulations. Common
gardens are grouped by the larger region they came from: North gardens
are within the natural range of the Midwest subpopulation, while Texas
gardens are within the natural range of the Gulf subpopulation.}

@@

\caption{\label{fig-qtl}Overlaps of QTL from an outbred pseudo-F2 cross
and with jointly re-estimated SNP effects in the 1\% tail of
significance from a diversity panel. Dotted lines indicate
permutation-based significance thresholds for each weather-related
function. Stars indicate QTL with significant enrichment for SNPs in the
1\% \emph{mash} tail; G, M, and B indicate which subpopulation had
enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both
subpopulations. Rug plots show genomic locations of SNPs in the 1\%
\emph{mash} tail for flowering date for each subpopulation. a) QTL
mapping for the start of vegetative growth (Greenup Date), and three
weather-related functions of greenup date. b) QTL mapping for the start
of reproductive growth (Flowering Date), and two weather-related
functions of flowering date.}

>>>>

Narrow-sense heritabilities (h\textsuperscript{2}) suggested that
rank-changing GxE for these phenotypes was present across the common
gardens (SI Appendix, Figure S1). h\textsuperscript{2} were typically
high at individual gardens: 59\% on average for green-up date, and 87\%
for flowering date. However, h\textsuperscript{2} were variable across
gardens, and green-up dates were uncorrelated (r\textsuperscript{2}
\textless{} 0.2) or negatively correlated between pairs of gardens
(Figure~\ref{fig-map} B). These negative and small correlations
undoubtedly contributed to the low h\textsuperscript{2} values for
green-up and flowering date when estimated jointly at all eight gardens:
h\textsuperscript{2} was 0.8\% for green-up and 23.2\% for flowering
date (SI Appendix, Figure S1).

@@

We expected patterns of GxE and GxWeather to vary at the locus,
subpopulation, and trait level; thus we were interested in modeling the
different ways effects might covary across our eight common gardens. We
jointly re-estimated genetic effects of a subset of SNPs across all
eight common gardens using \emph{mash} models, which can flexibly
capture many modes of covariance across gardens (SI Appendix, Sections
S1-S5, Datasets 1-6). To obtain genetic effects, we used subsets of
effects from garden- and subpopulation GWAS (SI Appendix, Section S2-S3;
Table S1). To obtain covariance structures, we specified both covariance
structures introduced in the initial \emph{mash} manuscript, and
GxWeather covariance matrices estimated from the covariance of empirical
weather patterns at each garden at specific times before the
phenological event (Table 1; SI Appendix, Section S1). Then, we used a
greedy algorithm to select covariance matrices from each model's set
that significantly improved the model likelihood (SI Appendix, Section
S4; Table S2). Finally, we fit \emph{mash} models using the selected set
of covariance matrices on a subset relatively unlinked
(r\textsuperscript{2} \textless{} 0.2), `strong' genetic effects with
low \emph{p}-values (SI Appendix, Section S5).

@@

Eight GxWeather covariance structures were selected by the greedy
algorithm, one to two per subpopulation: five for green-up date, and
three for flowering date (SI Appendix, Table S2). Of these eight, five
had mass on them in \emph{mash} models of the strong effects
(Figure~\ref{fig-covar} B,D). Three of these five matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while two had all positive or near-zero
covariances.

@@

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited rank-changing GxE, or a difference in sign
between pairs of Texas and North gardens (Figure~\ref{fig-effects} B; SI
Appendix, Fig S2). 78.7\% of pairwise comparisons between North and
Texas gardens had a difference in sign, while only 28.6\% and 0.2\% of
North-North or Texas-Texas comparisons had a difference in sign,
respectively (Figure~\ref{fig-effects} B). The majority of pairwise
effects for greenup for the Midwest (\textgreater55\%) and Both
(\textgreater85\%) subpopulations were the same sign, and effects most
often differed in magnitude between gardens within and between the
regions (Figure~\ref{fig-effects} B; SI Appendix, Fig S2).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited rank-changing GxE, or a difference in effect sign
within or between regions (Figure~\ref{fig-effects} B). More effects
differed in magnitude between the Texas and North regions than within
these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects} B),
while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering, but a large proportion
of these differed in sign between Texas and North regions (42.7\%) or
within the North region (65.4\%). Finally, in Both subpopulations, less
than 20\% of pairwise effects differed in sign (Figure~\ref{fig-effects}
B). Most differences in sign were between TX1, the southernmost garden,
and all other gardens (SI Appendix, Figure S3). Similarly, more effects
that differed in magnitude included gardens in the Texas region
(52.3-55.9\%), and most effect pairs in the North region were not
distinguishable (91.5\%).

@@

We sought additional experimental support the occurrence and location of
our significant SNP effect re-estimates using an independent pseudo-F2
mapping population created from Gulf \& Midwest individuals and grown at
the same sites (Figure~\ref{fig-qtl} A,B). We conducted quantitative
trait loci (QTL) mapping of flowering as functions of five environmental
cues that we also used as covariance matrices in \emph{mash}, and
identified eight QTL for flowering date, eight QTL for flowering as a
function of day length change two days prior, one QTL for the start of
vegetative growth, and two QTL for vegetative growth as a function of
daylength change one day prior, all of which showed QTL by environment
interactions (SI Appendix, Figure S4). All QTL for flowering overlapped
one or more homologs from rice or \emph{A. thaliana} with functionally
validated roles in flowering (SI Appendix, Dataset 7); the QTL on Chr04K
overlapped a gene recently functionally validated for flowering in
switchgrass (38). All flowering and green-up QTL intervals contained at
least one SNP significant in at least one \emph{mash} run at a
log10-transformed Bayes Factor \textgreater{} 2, or in the 1\% tail of
significance, whichever was stricter (SI Appendix, Dataset 8). We also
looked for enrichments of \emph{mash} SNPs in the 1\% tail of
significance (the `\emph{mash} 1\% tail') within each QTL interval. At
the 5\% level, three QTL had enrichments of SNPs in the \emph{mash} 1\%
tail. Overall, there were five significant enrichments (p \textless{}
0.05, hypergeometric test) of SNPs in the \emph{mash} 1\% tail in the
QTL intervals. Thus, we were able to experimentally support the genomic
windows of some re-estimates of significant SNP effects with a QTL
mapping experiment using a separate mapping population.

@@

\includegraphics{images/Figure_1_AB.png}

@@

\caption{\label{fig-map}Characterization of the timing of the onset of
vegetative (green-up) and reproductive (flowering) growth in the
switchgrass diversity panel. (a) Map and trait histograms of green-up
and flowering dates across two genetically distinct switchgrass
subpopulations and eight common gardens. Purple represents individuals
from the Midwest genetic subpopulation, and pink individuals from the
Gulf subpopulation; map positions represent the original collection
locations for the genotypes, and shapes represent the ecotype of the
genotype. Histogram vertical dashed lines indicate the summer solstice.
Common gardens are arranged in latitudinal order. (b) Phenotypic
correlations between clonal replicates planted at eight common gardens,
within and between two genetic subpopulations.}

@@

\includegraphics{images/Figure_2.png}

@@

\caption{\label{fig-covar}Example hypothesis-driven covariance matrices
specified in \emph{mash} and the posterior weights placed on all
covariance matrices. (a) Common gardens are arranged in latitudinal
order within the matrices. Top row: Four example canonical covariance
matrices. Canonical matrices (purple) have simple interpretations, such
as equal effects across all common gardens, or effects specific to a
single common garden. Bottom row: Five example GxWeather covariance
matrices specified for the green-up date or flowering date phenotype;
these matrices were created from environment-specific correlations
across eight common gardens, and are described in Table 1. (b,d) Total
posterior weight placed on each covariance matrix type specified for (b)
green-up date and (d) flowering date \emph{mash} models, within and
between two genetic subpopulations. Covariance matrices included in
\emph{mash} that had zero posterior weight in all three \emph{mash} runs
on the genetic subpopulations, such as the identity matrix, are not
shown. (c,e) Total posterior weight placed on covariance matrices that
were hypothesized or canonical, for the (c) green-up date phenotype and
(e) flowering date phenotype.}

@@

\caption{\label{fig-effects}Types of GxE present between of pairs of
jointly re-estimated effects in eight common gardens, for effects with
lfsr \textless{} 0.05. a) Examples of effect patterns at three pairs of
sites with three types of GxE. All effects are for an alternate allele,
with the reference allele effect defined to be zero at both gardens and
represented by the dashed vertical line. Sign: Effects that differ in
sign at these pairs of gardens (p \textless{} 0.05, lfsr). Magnitude:
Effects identical in sign (p \textless{} 0.05, lfsr) that differ in
magnitude by a factor of \textgreater0.4. Not Distinguishable: Effects
not distinguishable by magnitude nor sign of the effect, with no
measurable GxE. b) The fraction of effects with each GxE type for the
onset of vegetative growth (green-up date) and reproductive growth
(flowering date), within and between two genetic subpopulations. Common
gardens are grouped by the larger region they came from: North gardens
are within the natural range of the Midwest subpopulation, and include
MO, NE, MI, and SD, while Texas gardens are within the natural range of
the Gulf subpopulation, and include TX1, TX2, and TX3.}

@@

\caption{\label{fig-qtl}Overlaps of QTL from an outbred pseudo-F2 cross
and with jointly re-estimated SNP effects in the 1\% tail of
significance from a diversity panel. Dotted lines indicate
permutation-based significance thresholds for each weather-related
function. Stars indicate QTL with significant enrichment for SNPs in the
1\% \emph{mash} tail; G, M, and B indicate which subpopulation had
enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both
subpopulations. Rug plots show genomic locations of SNPs in the 1\%
\emph{mash} tail for flowering date for each subpopulation: Both
subpopulations are above the plot panel, the Gulf subpopulation is above
the x-axis, and the Midwest subpopulation is below the x-axis. a) QTL
mapping for the onset of vegetative growth (green-up date), and three
weather-related functions of green-up date. b) QTL mapping for the onset
of reproductive growth (flowering date), and two weather-related
functions of flowering date.}
